Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-09-09 supplement.tex
add Brenna's supp comments

Low coverage (4-8x) Illumina short read data were generated for the Nama,
Gumuz, Amhara and Oromo populations as part of the African Diversity Reference
Panel (Sanger / Wellcome Trust) \citep{Gurdasani2015-qy,Pagani2015-pz} and
approved through a secondary data analysis agreement for this project. Briefly,
raw reads were aligned to GRCH37 with BWAmem, duplicates
marked with Picard MarkDuplicates, reads were realigned around indels with GATK
RealignerTargetCreater/IndelRealigner followed by BQSR with dbSNP 137.
Contamination checks were performed requiring that FREEMIX $<0.05$;
contamination checks resulted in the elimination of 22 Nama samples. We note
that the high heterozygosity in these genomes both due to inherent genetic
diversity and admixture may have violated base assumptions for this
heterozygosity check. Genomes were then variant called with GATK3.2 Unified
Genotyper \citep{DePristo2011-up} using joint calling across 2,478 individuals within
the African Diversity Reference Panel (ADRP) dataset with a
minimum base quality of 17. Data were merged with 1000
Genomes Phase 3 \citep{1000_Genomes_Project_Consortium2015-zq} using the union
of sites identified with bcftools isec (-n+1) \citep{Danecek2021-kc} then refined
with Variant Recalibrator with a truth sensitivity threshold of 99.5\%. HapMap
III and dbSNP 138 served as known sites while 1000 Genomes Phase 1 Omni2.5 and
Phase 1 genomic SNP served as the training set. After VR, no batch effects were
observed along PC1 and PC2 for 1000 Genomes vs. ADRP. Phasing on the combined
dataset was performed via SHAPEIT2 \citep{Delaneau2013-aw} and utilized the duoHMM
option for duos and trios. We present new whole-genome data for 82 Nama individuals 
available under accession number EGAD00001006198.
Second- and third-degrees relatives were inferred from reconstructed
pedigrees with Omni2.5 SNP array data.
Among these 82 Nama samples, we down-sampled the individuals to minimize
close relatives and admixture, such that 44 Nama genomes were retained.
Individuals with $>70\%$ estimated Khoe-San ancestry were retained for
analysis, after partitioning ancestry into $K=6$ clusters with an initial
\texttt{ADMIXTURE} \citep{Alexander2009-sw} analysis on the full ADRP
dataset, where alternative ancestries represent European, West
African, Near Eastern, and East African gene flow.
Other \texttt{ADMIXTURE} analyses presented in this manuscript were
restricted to a subset of 1157 individuals representing a subset
of the full ADRP/Thousand Genomes dataset used in analyses presented here
(see Section~\ref{sec:populations-used}).

@@

DNA samples were collected from three Nama communities in the Richtersveld
region of South Africa, which borders southern Namibia, in 2012. A community
guide was present during each interview and facilitated consent in Afrikaans or
Nama. Written consent was recorded per our IRB protocol with human subjects
approval from Stanford University (Protocol \#13829), Stellenbosch University
(N11/07/210) and later maintained via SUNY Stony Brook (Protocol \#727494).
Saliva samples were extracted from Oragene [OGR-500] kits at Stellenbosch
University. Results stemming from genetic analyses have been communicated in
2015, 2019 and 2021 via community presentations, a radio interview and to
representatives of the Richtersveld National Park.

@@

For the demographic modeling, we further subset to the regional populations,
from West Africa: Mende from Sierra Leone (MSL); South Africa: Nama from South
Africa (newly sequenced); East Africa: Gumuz (traditionally hunter-gatherer
with low levels of Eurasian admixture), Oromo and Amhara (combined as
traditionally agriculturalists with large proportion of back-to-Africa Eurasian
ancestry); and Europe: British (GBR). These were combined with the
high-coverage Vindija Neanderthal genome \citep{Prufer2017-kk}, resulting in 290
total samples used in demographic inference.

@@

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{figures/supp-admixture-panels}
    \caption{
        \textbf{\texttt{ADMIXTURE} clustering using $K=2$ to $9$ clusters.}
        \texttt{ADMIXTURE} \citep{Alexander2009-sw} was used to cluster all
        individuals in the subset of populations used in analyses in this
        paper. At $K=5$ clusters, West African populations begin to show
        variable separation in assigned clusters, likely due to
        isolation-by-distance demographic processes rather than recent
        admixture events. Figure~\ref{fig:diversity}D in the main text shows
        the \texttt{ADMIXTURE} results for $K=4$, which most clearly displays
        known admixture events in the histories of East and South African
        populations.
    }
    \label{fig:supp-admixture}
\end{figure}
\end{document}

>>>>

Low coverage (4-8x) Illumina short read data were generated for the Nama,
Gumuz, Amhara and Oromo populations as part of the African Diversity Reference
Panel (Sanger / Wellcome Trust) \citep{Gurdasani2015-qy,Pagani2015-pz} and
approved through a secondary data analysis agreement for this project. Briefly,
raw reads were aligned to GRCH37 with BWAmem, duplicates
marked with Picard MarkDuplicates, reads were realigned around indels with GATK
RealignerTargetCreater/IndelRealigner followed by BQSR with dbSNP 137.
Contamination checks were performed requiring that FREEMIX $<0.05$;
contamination checks resulted in the elimination of 22 Nama samples. We note
that the high heterozygosity in these genomes both due to inherent genetic
diversity and admixture may have violated base assumptions for this
heterozygosity check. Genomes were then variant called with GATK3.2 Unified
Genotyper \citep{DePristo2011-up} using joint calling across 2,478 individuals within
the African Diversity Reference Panel (ADRP) dataset with a
minimum base quality of 17. Data were merged with 1000
Genomes Phase 3 \citep{1000_Genomes_Project_Consortium2015-zq} using the union
of sites identified with bcftools isec (-n+1) \citep{Danecek2021-kc} then refined
with Variant Recalibrator with a truth sensitivity threshold of 99.5\%. HapMap
III and dbSNP 138 served as known sites while 1000 Genomes Phase 1 Omni2.5 and
Phase 1 genomic SNP served as the training set. After VR, no batch effects were
observed along PC1 and PC2 for 1000 Genomes vs. ADRP. Phasing on the combined
dataset was performed via SHAPEIT2 \citep{Delaneau2013-aw} and utilized the duoHMM
option for duos and trios. We present new whole-genome data for 82 Nama individuals 
available under accession number EGAD00001006198.
Second- and third-degrees relatives were inferred from reconstructed
pedigrees with Omni2.5 SNP array data.
Among these 82 Nama samples, we down-sampled the individuals to minimize
close relatives and admixture, such that 44 Nama genomes were retained.

\subsubsection{ADMIXTURE} Nama individuals with $>70\%$ estimated Khoe-San
ancestry were retained for analysis here, after partitioning ancestry into $K =
6$ clusters with an \texttt{ADMIXTURE} \citep{Alexander2009-sw} analysis on the
full ADRP dataset \citep{Van_Eeden2022-od}. The 27 ADRP populations represent a
wide variety of West African, East African, North African and Bantu-speaking
South Africans. Europeans (1000 Genomes: CEU) were included to account for
colonial admixture. A Khoe-San specific ancestry emerged at $K = 3$,
predominant in the Nama. This ancestry remained stable at higher $K$'s. At $K =
6$, an ancestry representing the southern Bantu-speakers (Sotho and Zulu) also
emerged, at low frequency in some Nama. Higher $K$'s highlighted
population-specific ancestries or those with no representation in the Nama,
therefore we focused on $K = 6$ for selection of high Khoe-San ancestry
individuals. For illustration, \texttt{ADMIXTURE} analysis in
Figures~\ref{fig:diversity}D~and~\ref{fig:supp-admixture} was restricted to
1,157 individuals representing a subset of the full ADRP and several 1000
Genomes populations used in analyses presented here (see
Section~\ref{sec:populations-used}).

@@

\subsubsection{Ethics and Initiation of DNA Collection}

DNA samples were collected from three Nama communities in the Richtersveld
region of South Africa, which borders southern Namibia. Members of the Nama
community in the Richtersveld were initially approached regarding a genetic
ancestry study in 2011. The members included church elders, representatives at
the Richtersveld National Park including the appointee for Peoples and
Conservation (from which we obtained permission to sample individuals who still
live in the park), the Richtersveld Cultural World Heritage Site Coordinator,
and the Kuboes tourist office. All of these individuals self-identified as
Nama. There was strong support for research that examined the extent to which
the Nama are related to other Khoe-San populations, the impact of colonial
migration, the time depth of their presence in southern Africa and
relationships to other African groups. One primary concern was to make sure the
research results were reported back to the community. Based on the positive
response we received, permission was granted from the Stellenbosch University
HREC to continue the study. Written consent was recorded per our IRB protocol
with human subjects approval from Stanford University (Protocol \#13829),
Stellenbosch University (N11/07/210) and later maintained via SUNY Stony Brook
(Protocol \#727494). DNA were collected for the express purpose of understanding
Nama population history, their relationship to other Africans and human
evolution. No commercial activities are allowed. The IRB has approved the data
for public deposition and sharing provided that the research is related to
population history or human evolution.

\subsubsection{Collection Procedures}

Demographic and DNA collection was initiated in 2012 with a joint team of South
African and American researchers which included both geneticists and
anthropologists. Research assistants were fluent in Afrikaans, Nama and
English. Collection primarily occurred at home, by first approaching a family
member, gauging interest in the study, entering the home or sitting outside by
their invitation, oral and written consent occurring with family members
present and then finally -- completing the DNA sampling. By sampling
individuals at home, members of the family are able to voice concerns, whether
or not they decide to participate, and thus the final decisions are made in a
slow and ethical manner. This process sometimes involved introducing the study
and then returning at a later day in order to give participants sufficient time
to consider. Saliva samples were extracted from Oragene [OGR-500] kits at
Stellenbosch University. A copy of the English consent form is available as
a Supplementalal Document.

\subsubsection{Return of Results}

Results stemming from genetic analyses have been communicated in 2015, 2019 and
2021 via community presentations, a radio interview and in person to
representatives of the Richtersveld National Park. We created a poster
illustrating Khoe-San ancestry proportions, discriminating ancestry on maternal
vs. paternal lineages, the relationship between the Nama and Kalahari Khoe-San
populations, and the evolution of light skin pigmentation in southern Africa.
The poster was printed in Afrikaans and presented to the tourism office in
Kuboes (a central location) for display. A display panel on genetic ancestry is
currently under construction for the Richtersveld National Park entrance,
created in consultation with the authors here. Future community meetings are
anticipated in 2023. We remain committed to upholding this decade long
relationship with the Nama, and value the community’s input in the research
process.

%DNA samples were collected from three Nama communities in the Richtersveld
%region of South Africa, which borders southern Namibia, in 2012. A community
%guide was present during each interview and facilitated consent in Afrikaans or
%Nama. Written consent was recorded per our IRB protocol with human subjects
%approval from Stanford University (Protocol \#13829), Stellenbosch University
%(N11/07/210) and later maintained via SUNY Stony Brook (Protocol \#727494).
%Saliva samples were extracted from Oragene [OGR-500] kits at Stellenbosch
%University. Results stemming from genetic analyses have been communicated in
%2015, 2019 and 2021 via community presentations, a radio interview and to
%representatives of the Richtersveld National Park.

@@

For the demographic modeling, we further subset to the regional populations,
from West Africa: Mende from Sierra Leone (MSL); South Africa: Nama from South
Africa (newly sequenced, Figure~\ref{fig:supp-nama}); East Africa: Gumuz
(traditionally hunter-gatherer with low levels of Eurasian admixture), Oromo
and Amhara (combined as traditionally agriculturalists with large proportion of
back-to-Africa Eurasian ancestry); and Europe: British (GBR). These were
combined with the high-coverage Vindija Neanderthal genome
\citep{Prufer2017-kk}, resulting in 290 total samples used in demographic
inference.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/SuppFig_HDV.60_CEU.Nama}
    \caption{caption text}
    \label{fig:supp-nama}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[width=5.75in]{figures/supp-admixture-panels}
    \caption{
        \textbf{\texttt{ADMIXTURE} clustering using $K=2$ to $9$ clusters.}
        \texttt{ADMIXTURE} \citep{Alexander2009-sw} was used to cluster all
        individuals in the subset of populations used in analyses in this
        paper. At $K=5$ clusters, West African populations begin to show
        variable separation in assigned clusters, likely due to
        isolation-by-distance demographic processes rather than recent
        admixture events. Figure~\ref{fig:diversity}D in the main text shows
        the \texttt{ADMIXTURE} results for $K=4$, which most clearly displays
        known admixture events in the histories of East and South African
        populations.
    }
    \label{fig:supp-admixture}
\end{figure}

@@

\end{document}
